---
title: "Final Paper"
author: "Jake Becker, Evan Kessler, Ali Almarbooei"
date: "11/24/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(MASS)
library(caret)
library(class)
library(tree)
bobsled <- read.csv("data.csv") %>%
  mutate(Medal = factor(Medal, levels = c("No Medal", "Bronze", "Silver", "Gold", ordered = TRUE)))
set.seed(445)

```
# Project Motivation

  Our project is largely motivated by the documentary Icarus, which focuses on the use of performance enhancing drugs (PEDs) in professional sports, and the Russian doping scandal following the 2014 Winter Olympics. After the 2014 winter Olympics word got out about the Russian team doping, specifically in the winter Olympics Our assumption was this was the case because they were hosting these games in Sochi and they wanted to appeal to their home crowd. After some research on the games there were no real determinations on who actually would have one the medals in the sports where medalers were caught doping. We wanted to take the opportunity to try and predict the outcome of the 2014 Winter Olympics, with the information we could obtain. Unfortunately, most of the sports that are bracket based, where predicting winners makes sense since strong teams can be eliminated early by participants that are cheating, are largely team based in the winter olympics. The only athletes involved in the scandal who had strong enough performances to reevaluate were the members of the men's two and four man bobsled teams or the cross-country skiing. After deliberating on which sport we thought would be best we ended on bobsleighing as it seemed more interesting and the medalers were all gold. This worked well because all members of the team were involved, and teams are small enough where individual contributions to results are still relevant.  

# Data

Part of the reason we decided on looking at olympic data was because we found a dataset on Kaggle.com. The dataset was titled, "120 years of Olympic history: athletes and results" and included every olympic athlete that participated in either winter or summer games between 1896-2016. This ended being a massive dataset clockign in at around 300,000 rows (athletes) and 15 columns. Sorting through this we ended up having to do a few things. First, even though these games are national events and each athlete has to provide data on their height, weight, and age there were a few athletes, even recently, that did not contain their height, weight, or age. To combat this we decided to take an average for that sport and apply it to the NA values corresponding to said athletes. Therefore, some of our data is manipulated, although we figured it would be best to include them anyways as an average weight/height/age would not be too far off form the real thing. The second thing was selecting all of the athletes that participated in bobsledding. Since all of the bobsled athletes that were caught doping and medaled were male we decided to only focus on male two man and four man bobsledding. Initially we were looking at datasets with both male and female athletes. This proved challenging because there ended up being two chunks of athletes, one being female weight and height and the other being male. We found that some of the models struggled to predict solely because there was such a split between the two genders data. Out of the remaining mens bobsledding athletes there were a total of 10 of the 43 confirmed dopers throughout the games. 6 gold medals were received from those 10 athletes, two being from the tandem of the two-man bobsleigh competition and four from the four-man. Lastly it was up to the manipulation of the medal winners. The way the kaggle dataset depicted medals was either Gold, Silver, Bronze, or NA, to combat this we ended up factoring each of these, first being No Medal then Bronze, Silver, and finally Gold. In addition we manipulated the dataset to add in a final column titled 'MedalBin' which held a value of 1 if the athlete medaled and 0 if they did not. We found that for some of our models it was best to use a binary variable as our response. After all of this manipulation we went from 300,000 athletes to just 710 from the games in 2002 to 2014. For most of our models we used training and testing datasets, because our goal of this project was to try and predict the true winneres of the 2014 olympics instead of using traditional training and testing data with random samples where around 70% of the data is training and the remaining is testing; we decided to just look at some of the most recent olympic games, 2002-2010 and set these as our training while the sole year of 2014 was our testing. 

# Trees, Bagging, and Random Forests

One of the first models we decided to attempt to implement was trees. We figured this would be a great start as the variables we were using were already sort of boxed off due to lower limits on bobsledders based on typical weight and height. 

```{r}
bobsledT = bobsled %>%
  filter(Year != 2014)
bobsled14 = bobsled %>%
  filter(Year == 2014)
bobsled14 = na.omit(bobsled14)

dftree <- tree(Medal ~ Height + Weight + Age + NOC + MedalBin, data = bobsledT)
# Gives us Weight, Height, Age, Year, X (ID) as best predictors

plot(dftree); text(dftree)

confusion <- table(pred = predict(dftree, bobsled14, type = "class"), true = bobsled14$Medal)
confusion
```

As you can see from our confusion matrix doing basic trees we did not receive very good results. It did an excellent job at prediction whether an athlete did not medal due to the fact that it utilized our column of MedalBin, meaning anything under .5 (see tree) in that column would equal not medaling. This ended up giving us an error rate for No Medal of 0.00%, amazing but not entirely important as it seems that it only predicted 5 out of 18 medals correctly between Bronze and Gold. 

# Unsupervised Learning    

  In order to learn more about the data, we processed the reduced year set through a hierarchical clustering algorithm.  This method was used over k-means clustering or Principal Component Analysis due to the heavy use of categorical variables. This was unfortunate as our data was very high dimensional, and use of PCA to help reduce this would have been useful. This applies as well to Principal Component Regression.
Hierarchical clustering pruned to 4 clusters lead to interesting results, and a large amount of participants in past years medaling placed into clusters 3 and 4.


```{r unsupervised, echo = FALSE}
d = dist(bobsled, method = 'euclidean')
clust = hclust(d, method = 'complete')
prunclust = cutree(clust, k = 4)
plot(clust)
```

```{r clusterplot, echo = FALSE}
df = bobsled %>%
  mutate(cluster = as.character(prunclust))
df %>%
  ggplot(aes(x = Weight,
             y = Height,
             col = cluster)) + 
  geom_point()
```


```{r medalplot, echo = FALSE}
bobsled %>%
  ggplot() + 
  geom_point(aes(x = Weight,
                 y = Height,
                 col = Medal))
```

```{r clustcount, echo = FALSE}
df %>%
  group_by(cluster) %>%
  summarise(Gold = sum(Medal == 'Gold'),
            Silver = sum(Medal == 'Silver'),
            Bronze = sum(Medal == 'Bronze'),
            DNM = sum(Medal == 'No Medal'))
```

```{r traintest, include = FALSE}
bobsledT = bobsled %>%
  filter(Year != 2014)
bobsled14 = bobsled %>%
  filter(Year == 2014)
bobsled14 = na.omit(bobsled14)
```

# K Nearest Neighbors

  We initially attempted to fit a K Nearest Neighbors classification model to predict medal winners.  This quickly proved to be an ineffective and inappropriate choice of algorithm.  First of all, there is no way to guarantee the model predicts a medal (Pictured below), and the calculation method for KNN does not give probability for being in a certain class.  The second problem being the binary response, as this relates back to the problem of an ordinal response.  We quickly moved past this method to focus on ordinal logistic regression.

```{r knntraintest, echo = FALSE}
bobknn = knn3(MedalBin ~ Age + Height + Weight, data = bobsledT, k = 3)
bobpred = predict(bobknn, newdata = bobsled14)
bobpredvec = c(ifelse(bobpred[,1] > .5, 0, 1))
confusion = table(actual = bobsled14$MedalBin, predicted = bobpredvec)
confusion
```

```{r knnplot, echo = FALSE}
plotframe = data.frame(bobsled14, predicted = bobpredvec)
ggplot() + 
  geom_point(aes(x = Weight,
                 y = Height,
                 col = predicted),
             data = plotframe)
```

# Ordinal Logistic Regression  

  Shortly after we moved to ordinal logistic regression which proved to be much more successful than other methods.  Ordinal logistic regression returns log odds of placement into multiple ordered factor variables.  This algorithm is most useful with problems that include a ordered response.  Since our response variable is ordered, it seemed much more appropriate. This did require reshaping our response variable to be an ordered factor. 
  Initially we struggled to predict results with the model, due to small testing data size, and countries making their first entrance into the contest in years.  Other problems included NAs and computational issues.  However, we were able to fit a ordinal logistic regression model with Age, Weight, Height, Event (two or four man), and NOC as predictors.  We would have liked to use team, but that would have lead to many more problems similar to that with the Austria country code. 

```{r probit model, echo=FALSE}
probit = polr(as.factor(Medal) ~ Age + Weight + Height + Event + NOC, data = bobsledT, Hess = TRUE)
probpred = predict(probit, newdata = bobsled14, type = 'probs')
summary(probit)
probitDf = cbind(bobsled14, probpred)
```

  For each participant, we aggregated their probability to win each medal, and averaged the value with that of their teammates.  Each team's average medal win probability was used to predict the winners.

```{r mensfourpred, echo = FALSE}
probitDf %>% 
  filter(Event == "Bobsleigh Men's Four") %>%
  mutate(MedalProb = Gold + Silver + Bronze) %>% 
  group_by(Team) %>%
  summarize(Win = round(mean(MedalProb), digits = 4)) %>%
  arrange(desc(Win))
```

True 4 man results   
1. Russia 1 (DQ'd)  
2. Latvia 1  
3. US 1  
4. Russia 2  
5. GB 1  
6. Germany 1  
7. Germany 3    
8. Switzerland 1  
9. Canada 2  
10. Germany 2  
  
Looking at our results for the top 10 in four man bobsledding, we can see that there were only two predictions outside of the original winnings top 10 of which we predicted. Due to the fact that our predicitons are based off of weight, height, age, NOC, and event, there will be some natural differences in placements. This would be a good estimation on an upcoming olympics based on initial weigh in data to predict who could win. Otherwise, we have predicted that the russian doping team is still in the top 10 however, without the advantage they had in the real games with doping, they are only predicited to get 8th place. This is fantastic news as this means their team when looking at usual placements in our training data should not have preformed as well as they did. 

```{r menstwopred, echo = FALSE}
probitDf %>% 
  filter(Event == "Bobsleigh Men's Two") %>%
  mutate(MedalProb = Gold + Silver + Bronze) %>% 
  group_by(Team) %>%
  summarize(Win = round(mean(MedalProb), digits = 4)) %>%
  arrange(desc(Win))
```

True 2 man results:  
1. Russia 1 (DQ)  
2. Switzerland 1  
3. US 1  
4. Russia 2  
5. Latvia 1  
6. Canada 3  
7. Canada 2  
8. Germany 1  
9. Canada 1  
10. Switzerland 2  

For the two man bobsledding we see similar results as four man, only a couple of teams were predicted that were not already included in the top 10 of the games. These extras actually ended up placing just outside of the top ten so the predictions aren't incredibly outlandish. Just like in four man we see that the russia team that was caught doping and medaling gold (Russia 1) placed extremely low, 9th, meaning they should not have been slated to win the gold. 

This corresponds greatly with the documentary Iccarus that initially motivated us into looking into the 2014 Sochi games. The documentarian ended up doping himself for a year of cycling to see the general affects of the drugs. What he found after a year was his height and weight really didn't defer from the year prior in which he wasn't doping. All that changed was his stamina and muscle fatigue. Knowing of this, the heights and weights of the russian athletes would not have seen a change post-doping meaning that the real advantage would not be able to be measured. 

```{r resultsfour, echo = FALSE}
probitDf %>% 
  filter(Event == "Bobsleigh Men's Four") %>%
  mutate(MedalProb = Gold + Silver + Bronze) %>% 
  ggplot() + 
  geom_point(aes(x = Weight,
                 y = Height,
                 col = MedalProb)) + 
  ggtitle("Men's Fours")
```

```{r twomanresults, echo = FALSE}
probitDf %>% 
  filter(Event == "Bobsleigh Men's Two") %>%
  mutate(MedalProb = Gold + Silver + Bronze) %>% 
  ggplot() + 
  geom_point(aes(x = Weight,
                 y = Height,
                 col = MedalProb)) + 
  ggtitle("Men's Twos")
```

```{r allmens, echo = FALSE}
probitDf %>%
  mutate(MedalProb = Gold + Silver + Bronze) %>%
  ggplot() + 
  geom_point(aes(x = Weight,
                 y = Height,
                 col = MedalProb)) + 
  ggtitle("All Men's")
df %>%
  filter(Year == 2014) %>%
  ggplot(aes(x = Weight,
             y = Height,
             col = cluster)) + 
  geom_point() + 
  ggtitle("Clusters")
```

  As you can see above, many of the predicted medal winners were in clusters 3 and 4 still, suggesting some interaction between demographics and country to predict medal winners.

Works Cited:
Jones, Matt. “Bobsled Medal Results and Times from Olympic 2014 Men's 4-Man.” Bleacher Report, Bleacher Report, 1 Oct. 2017, https://bleacherreport.com/articles/1970485-bobsled-medal-results-and-times-from-olympic-2014-mens-four-man. 

Wood, Donald. “Bobsled Medal Results and Times from 2014 Olympics Men's 2-Man Event.” Bleacher Report, Bleacher Report, 22 Aug. 2017, https://bleacherreport.com/articles/1963113-bobsled-medal-results-and-times-from-olympic-2014-mens-2-man-event#:~:text=2014%202-Man%20Bobsled%20Top%2010%20Results%20%20,%20%20%2B1.57%20%208%20more%20rows%20. 

Analytics, Perceptive. “How to Perform Ordinal Logistic Regression in R: R-Bloggers.” R, 18 June 2019, https://www.r-bloggers.com/2019/06/how-to-perform-ordinal-logistic-regression-in-r/. 
